\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-11 analysis/paper/paper.tex
edit first case study

To introduce the different features available, we make use of the
European Randomized Study of Prostate Cancer Screening data; this
dataset is available through the \pkg{casebase} package:

@@

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R>                        levels = c(0,1), 
R>                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

The results of this study were published by {[}schroder2009screening{]}.
This data was obtained using the approach described in
{[}liu2014recovering{]}.

Population time plots can be extremely informative graphical displays of
survival data. They should be the first step in an exploratory data
analysis. We facilitate this task in the \pkg{casebase} package using
the \code{popTime} function. We first create the necessary dataset for
producing the population time plots, and we can produce the plot by
using the corresponding \code{plot} method:

@@

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}
\end{CodeChunk}

First, we fit a Cox model to the data, examine the hazard ratio for the
screening group (relative to the control group), and plot the cumulative
incidence function (CIF).

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R>                              data = ERSPC)
R> summary(cox_model)
\end{CodeInput}
\end{CodeChunk}

We can plot the CIF for each group:

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R>                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]))
R> legend("topleft", 
R>        legend = c("Control group", "Screening group"), 
R>        col = c("red","blue"),
R>        lty = c(1, 1), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

Next we fit several models using case-base sampling. The models we fit
differ in how we choose to model time.

The \code{fitSmoothHazard} function provides an estimate of the hazard
function \(h(x, t)\) is the hazard function, where \(t\) denotes the
numerical value of a point in prognostic/prospective time and \(x\) is
the realization of the vector \(X\) of variates based on the patient's
profile and intervention (if any).

@@

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
R>                                                   data = ERSPC, 
R>                                                   ratio = 100)
R> 
R> summary(casebase_exponential)
R> exp(coef(casebase_exponential)[2])
R> exp(confint(casebase_exponential)[2,])
\end{CodeInput}
\end{CodeChunk}

The \code{absoluteRisk} function provides an estimate of the cumulative
incidence curves for a specific risk profile using the following
equation:

\[ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]

In the plot below, we overlay the estimated CIF from the casebase
exponential model on the Cox model CIF:

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R>                                           time = seq(0,15,0.1), 
R>                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

As we can see, the exponential model is not a good fit. Based on what we
observed in the population time plot, where more events are observed
later on in time, this poor fit is expected. A constant hazard model
would overestimate the cumulative incidence earlier on in time, and
underestimate it later on; this is what we see on the cumulative
incidence plot. This example demonstrates the benefits of population
time plots as an exploratory analysis tool.

Next we enter time linearly into the model:

@@

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
R>                                  data = ERSPC, 
R>                                  ratio = 100)
R> 
R> summary(casebase_time)
R> exp(coef(casebase_time))
R> exp(confint(casebase_time))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R>                                           time = seq(0,15,0.1), 
R>                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

We see that the Weibull model leads to a better fit.

Next we try to enter a smooth function of time into the model using the
\code{splines} package:

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
R>                                     data = ERSPC, 
R>                                     ratio = 100)
R> 
R> summary(casebase_splines)
R> exp(coef(casebase_splines))
R> exp(confint(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R>                                     time = seq(0,15,0.1), 
R>                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

It looks like the best fit.

Since we are within the GLM framework, we can easily test for which
model better fits the data using a Likelihood Ratio Test (LRT). The null
hypothesis here is that the linear model is just as good as the larger
(in terms of number of parameters) splines model.

@@

\begin{CodeInput}
R> anova(casebase_time, casebase_splines, test = "LRT")
\end{CodeInput}
\end{CodeChunk}

As expected, we see that splines model provides a better fit.

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    798    759    737    708
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>         
#> 365 0.32
\end{CodeOutput}

@@

\begin{CodeOutput}
#> [1] 0.24
\end{CodeOutput}
\end{CodeChunk}

@@

This report was generated on 2019-12-11 10:31:00 using the following
computational environment and dependencies:

@@

@@

\begin{CodeOutput}
#> Local:    review-glmnet /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [b7e23a6] 2019-12-10: Merge branch 'master' into review-glmnet
\end{CodeOutput}
\end{CodeChunk}

>>>>

For the first case study, we will use of the European Randomized Study
of Prostate Cancer Screening data. This dataset is available through the
\pkg{casebase} package:

@@

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R+                        levels = c(0, 1), 
R+                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

The results of this study were published by
\citep{schroder2009screening}, and the dataset itself was obtained using
the approach described in \citep{liu2014recovering}.

Population time plots can be extremely informative graphical displays of
survival data. They should be the first step in an exploratory data
analysis. We facilitate this task in the \pkg{casebase} package using
the \code{popTime} function. We first create the necessary dataset for
producing the population time plots, and we can generate the plot by
using the corresponding \code{plot} method:

@@

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}

\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}

ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.

Next, we investigate the differences between the control and the
screening arms. A common choice for this type of analysis is to use a
Cox regression model and estimate the hazard ratio for the screening
group (relative to the control group). In \proglang{R}, it can be done
as follows:

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R+                              data = ERSPC)
R> summary(cox_model)
\end{CodeInput}

\begin{CodeOutput}
#> Call:
#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
#>     ScrArm, data = ERSPC)
#> 
#>   n= 159893, number of events= 540 
#> 
#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#>                       exp(coef) exp(-coef) lower .95 upper .95
#> ScrArmScreening group     0.801       1.25     0.674     0.952
#> 
#> Concordance= 0.519  (se = 0.011 )
#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
#> Wald test            = 6.37  on 1 df,   p=0.01
#> Score (logrank) test = 6.39  on 1 df,   p=0.01
\end{CodeOutput}
\end{CodeChunk}

We can also plot the cumulative incidence function (CIF) for each group.
However, this involves estimating the baseline hazard, which Cox
regression treated as a nuisance parameter.

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R+                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> legend("topleft", 
R+        legend = c("Control group", "Screening group"), 
R+        col = c("red","blue"),
R+        lty = c(1, 1), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}

Next, we show how case-base sampling can be used to carry out a similar
analysis. We fit several models that differ in how we choose to model
time.

First, we will fit an exponential model. Recall that this corresponds to
excluding time from the linear predictor.

@@

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
R+                                                   data = ERSPC, 
R+                                                   ratio = 100)
R> 
R> summary(casebase_exponential)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.148  -0.148  -0.148  -0.133   3.079  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value            Pr(>|z|)    
#> (Intercept)            -7.7650     0.0557 -139.44 <0.0000000000000002 ***
#> ScrArmScreening group  -0.2166     0.0884   -2.45               0.014 *  
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 6052.9  on 54538  degrees of freedom
#> AIC: 6057
#> 
#> Number of Fisher Scoring iterations: 7
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_exponential)[2])
\end{CodeInput}

\begin{CodeOutput}
#> ScrArmScreening group 
#>                  0.81
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_exponential)[2,])
\end{CodeInput}

\begin{CodeOutput}
#>  2.5 % 97.5 % 
#>   0.68   0.96
\end{CodeOutput}
\end{CodeChunk}

We can then use the \code{absoluteRisk} function to get an estimate of
the cumulative incidence curve for a specific covariate profile. In the
plot below, we overlay the estimated CIF from the exponential model on
the Cox model CIF:

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}

As we can see, the exponential model gives us an estimate of the hazard
ratio that is similar to the one obtained using Cox regression. However,
the CIF estimates are quite different. Based on what we observed in the
population time plot, where more events are observed later on in time,
we do not expect a constant hazard would be a good description for this
data. In other words, this poor fit for the exponential hazard is
expected. A constant hazard model would overestimate the cumulative
incidence earlier on in time, and underestimate it later on; this is
what we see on the cumulative incidence plot. This example demonstrates
the benefits of population time plots as an exploratory analysis tool.

For the next model, we include time as a linear term. Recall that this
corresponds to a Weibull model.

@@

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
R+                                  data = ERSPC, 
R+                                  ratio = 100)
R> 
R> summary(casebase_time)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.389  -0.161  -0.122  -0.095   3.464  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value            Pr(>|z|)    
#> (Intercept)            -9.0257     0.1122  -80.42 <0.0000000000000002 ***
#> Follow.Up.Time          0.2207     0.0145   15.18 <0.0000000000000002 ***
#> ScrArmScreening group  -0.2311     0.0886   -2.61              0.0091 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059  on 54539  degrees of freedom
#> Residual deviance: 5815  on 54537  degrees of freedom
#> AIC: 5821
#> 
#> Number of Fisher Scoring iterations: 8
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
#>               0.00012               1.24701               0.79370
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
#>                          2.5 %  97.5 %
#> (Intercept)           0.000096 0.00015
#> Follow.Up.Time        1.212086 1.28321
#> ScrArmScreening group 0.666385 0.94341
\end{CodeOutput}
\end{CodeChunk}

Again, the estimate of the hazard ratio is similar to that obtained from
Cox regression. We then look at the estimate of the CIF:

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}

We see that the Weibull model leads to a better fit of the CIF than the
Exponential model, when compared with the semi-parametric approach.

Finally, we can model the hazard as a smooth function of time using the
\code{splines} package:

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
R+                                     data = ERSPC, 
R+                                     ratio = 100)
R> 
R> summary(casebase_splines)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.397  -0.173  -0.136  -0.085   3.797  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value             Pr(>|z|)    
#> (Intercept)           -10.2787     0.3175  -32.38 < 0.0000000000000002 ***
#> bs(Follow.Up.Time)1     4.3191     0.8062    5.36        0.00000008456 ***
#> bs(Follow.Up.Time)2     2.1061     0.4878    4.32        0.00001580161 ***
#> bs(Follow.Up.Time)3     4.5310     0.7160    6.33        0.00000000025 ***
#> ScrArmScreening group  -0.2068     0.0886   -2.33                 0.02 *  
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5792.2  on 54535  degrees of freedom
#> AIC: 5802
#> 
#> Number of Fisher Scoring iterations: 9
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
#>              0.000034             75.119228              8.216108 
#>   bs(Follow.Up.Time)3 ScrArmScreening group 
#>             92.847641              0.813203
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
#>                           2.5 %     97.5 %
#> (Intercept)            0.000018   0.000062
#> bs(Follow.Up.Time)1   15.997506 378.737706
#> bs(Follow.Up.Time)2    3.256838  22.143147
#> bs(Follow.Up.Time)3   21.638298 362.098155
#> ScrArmScreening group  0.682786   0.966562
\end{CodeOutput}
\end{CodeChunk}

Once again, we can see that the estimate of the hazard ratio is similar
to that from the Cox regression. We then look at the estimate of the
CIF:

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R+                                     time = seq(0,15,0.1), 
R+                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}

Qualitatively, the combination of splines and case-base sampling
produces the parametric model that most closely resembles the Cox model.
In the following table, we see that the confidence intervals are also
similar across all four models. In other words, this reinforces the idea
that, under proportional hazards, we do not need to model the full
hazard to obtain reliable estimates of the hazard ratio. Of course,
different parametric models for the hazards give rise to qualitatively
different estimates for the CIF.

@@

\begin{tabular}{l|r|l|r}
\hline
Model & HR & CI & Width\\
\hline
Cox & 0.80 & (0.67, 0.95) & 0.28\\
\hline
Exponential & 0.81 & (0.68, 0.96) & 0.28\\
\hline
Weibull & 0.79 & (0.67, 0.94) & 0.28\\
\hline
Splines & 0.81 & (0.68, 0.97) & 0.28\\
\hline
\end{tabular}

As noted above, the usual asymptotic results hold for likelihood ratio
tests built using case-base sampling models. Therefore, we can easily
test the null hypothesis that the exponential model is just as good as
the larger (in terms of number of parameters) splines model.

\begin{CodeInput}
R> anova(casebase_exponential, casebase_splines, test = "LRT")
\end{CodeInput}

\begin{CodeOutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
#> 1     54538       6053                                    
#> 2     54535       5792  3      261 <0.0000000000000002 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

As expected, we see that splines model provides a better fit. Similarly,
we can compare the AIC for all three case-base models:

\begin{CodeInput}
R> AIC(casebase_exponential)
\end{CodeInput}

\begin{CodeOutput}
#> [1] 6057
\end{CodeOutput}

\begin{CodeInput}
R> AIC(casebase_time)
\end{CodeInput}

\begin{CodeOutput}
#> [1] 5821
\end{CodeOutput}

\begin{CodeInput}
R> AIC(casebase_splines)
\end{CodeInput}

\begin{CodeOutput}
#> [1] 5802
\end{CodeOutput}
\end{CodeChunk}

Once again, we have evidence that the splines model provides the best
fit.

With this first case study, we saw how population-time plots can provide
a useful summary of the incidence for our event of interests. In
particular, we saw how it can provide evidence against an exponential
hazard model. We then explored how different parametric models can
provide different estimates of the CIF, even though they all give
similar estimates for the hazard ratios.

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    810    772    741    722
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>         
#> 365 0.25
\end{CodeOutput}

@@

\begin{CodeOutput}
#> [1] 0.2
\end{CodeOutput}
\end{CodeChunk}

@@

This report was generated on 2019-12-11 15:19:04 using the following
computational environment and dependencies:

@@

#>  MASS          7.3-51.4   2019-04-26 [4]

@@

\begin{CodeOutput}
#> Local:    review-glmnet /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [491ef55] 2019-12-11: edit third case study
\end{CodeOutput}
\end{CodeChunk}
